---
title: Mini Data Analysis
Author: Tigana Runte
output: github_document
---

# Task 1


## Choose 4 data sets
After an initial pass I have selected the following data sets to explore in more detail:

1: `flow_sample`
2: `vancouver_trees`
3: `cancer_sample`
4: `steam_games`

## Bellow I explore these 3 attributes of the data sets from the datateacher package.

First I load needed packages.
```{r}
#install.packages("devtools")
#devtools::install_github("UBC-MDS/datateachr")
```
```{r}
#install.packages("datateachr")
library(datateachr)
library(tidyverse)
```
### Now I look at some attributes of the data set `flow_sample`.

```{r}
glimpse(flow_sample)

```
Three attributes of `flow_sample`:

1) It has 7 columns and 218 rows. Please note I later removed the column `station_id` as there was only the one value. Unfortunately, this had the unintended consequence of removing the variable up here where I glimpse the data as well. 

2) Has both quantitative and qualitative variables. 

3) There appears to be quite a few missing values in this data set.

### Next we look at the data set `vancouver_trees`.


```{r}
glimpse(vancouver_trees)

```
Three attributes of `vancouver_trees`:

1) It has 20 columns and 146,611 rows.

2) Has quantitative (<dbl>) and qualitative (<chr>) variables. As well as a date variable( <date>).

3) The qualitative variables appear to mostly be categorical. eg: `genus_name`.

###Next we look at the data set `cancer_sample`

```{r}
glimpse(cancer_sample)

```
Three attributes of `cancer_sample`:

1) There are 32 columns and 569 rows.

2) There is only one qualitative variable `diagnosis`, the rest are quantitative.

3) Except for the variable `diagnosis`, all variables appear to have an associated standard error. For example we have the variables `area_mean` and `area_se`

### Finally we explore the `steam_games` data set

```{r}
glimpse(steam_games)

```
Three attributes of `steam_games`:
1) There are 21 columns and 40,833 rows.

2)There are both quantitative and qualitative variables.

3) There are multiple qualitative variables that take free form text.


## Top 2 data sets:
I am most interested in the data sets `flow_sample` and `cancer_sample`.

I am interested in the `cancer_sample` data set for personal reasons. My father was recently diagnosed with cancer and it has cause me to become very interested in cancer research. 

I am interested in the `flow_sample` data set as it has multiple missing values and I am interested to see what effect this might have on the analysis. Also I lived in Banff for a while, and spent lots of time on the bow river. 

## Top Choice: `Flow_sample`.

I eventually settled on `flow_sample` as my main data set. There are only 7 (6 because the 7th is in the void) columns which I believe will make it easier to give each variable attention in its relationship to flow. There are also multiple missing values and I want to explore different ways to work with this.

# Task 2
## This is my main analysis.
```{r}
head(flow_sample)
```

 
### 1) The first thing I Want to do is to Mutate a variable. 
```{r}
table(flow_sample$sym)
sum(is.na(flow_sample$sym))
```
Most of the values for sym are "B". There are 119 missing values and 2 values for "A" and "E" respectively. I am going to mutate this column into the categories B and other. I also removed the column `station_id` as discussed above.
```{r}
flow_sample <- flow_sample %>% 
  #select(-station_id) %>% 
  mutate(sym = ifelse(is.na(sym), "Other",
                      ifelse(sym == "B", "B", "Other")))

table(flow_sample$sym)
head(flow_sample)
```


### 2)Plot distribution of flow
```{r}
ggplot(flow_sample) +
  aes(x=flow) +
  geom_histogram() +
  labs(
    title = "Frequency distribution of flow",
    x = "Flow",
    y = "Absolute frequency"
  )
```
It appears `flow` takes on a large range of values. There is a large concentration of flow rates under 10 m^3/s, while the rest of the values seem to be concentrated between 100 and 400 m^3/s with an outlier of 466 (flood of 2013). This potentialy represents the split `minimum` and `maximum` values for `extreme_type`.

```{r}
sum(is.na(flow_sample$flow))

```
I also quickly checked to see if `flow` has any missing values. There are only two. 

### 3) flow vs year

```{r}
flow_sample %>% 
  group_by(year) %>% 
  summarise(cumflow = sum(flow, na.rm = T)) %>% 
  ggplot() +
  aes(x=year, y=cumflow) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Flow across the years",
    x = "Year",
    y = "Flow"
  )  
```
We can see from the plot that the overall `flow` has decreased slightly over the years, with a notable peak in 2013 when we know there was a flood. However, on the shorter scale, (year to year), there is a lot of variation. I would like to further explore why this is.

### 4) flow of a particular month

I would like to look at the flow rate by `month`. Perhaps the differences year to year can be partially explained by the month in which the data was collected.

```{r}
flow_sample %>% 
  group_by(month) %>% 
  summarise(cumflow = sum(flow, na.rm = T)) %>% 
  ggplot() +
  aes(x=month, y=cumflow) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = unique(flow_sample$month)) +
  labs(
    title = "Cumulative flow along the year",
    caption = "Period: 1909 to 2018",
    x = "Month",
    y = "Flow"
  ) 
```
There seems to be a strong relationship between `month` and `flow`. Cumulative flow is very low from month 1 to 4, and, 8-12. It seems likely this can be explained by the seasons in which the flow was collected, with the warmer season having much higher flow rates.  I just also want to note that there are no observations for months 9 and 10. 


### 5) Boxplot frequency `B` vs `Other`

```{r}
ggplot(flow_sample) +
  aes(x=sym, y=flow, fill=sym) +
  geom_boxplot() +
  facet_wrap(~sym, scales = "free") +
  labs(
    title = "Boxplot of flow by sym",
    x = "Sym",
    y = "Flow"
  ) +
  theme(legend.position = "none") 
```
There are a few things to note here. We can see that all flow values associated with `B` are very low with a small IQR and only 1 outlier. The flow values associated with `Other` are much higher with a much larger IQR and more outliers. This tells us that flow rates that were collected in icey conditions `B`, are much lower than those collected at other times. However there are 119 missing values and I worry this will effect the interpretation. I would like to approximate the effect of icy vs not icy conditions using a new variable seasons. This should approximate the weather conditions that they were collected in while reducing the issue of the missing values.



### 6) Boxplot frequency `seasons`
First I make the new variable `season` to show what season the sample was collected in. 
```{r}

flow_sample <- flow_sample %>% 
  mutate(season = case_when(
    month > 5 & month < 9 ~ "Summer",
    month > 2 & month < 6 ~ "Spring",
    (month > 0 & month < 3) | month == 12  ~ "Winter",
    month > 8 & month < 12 ~ "Autumn",
    TRUE ~ as.character(month)
  ) %>% as.factor()
  )

```

Next I created boxplots for flow based on season.
```{r}
flow_sample <- flow_sample %>% 
  mutate(Season_Flow = ifelse(season == "Spring" | season == "Summer", "High Flow", "Low Flow"))

flow_sample %>% 
  filter(!is.na(season)) %>% 
  ggplot() +
  aes(x=season, y=flow, fill= season) +
  geom_boxplot() +
  facet_wrap(~Season_Flow, scales = "free") +
  labs(
    title = "Boxplot of flow by Season",
    x = "Season",
    y = "Flow"
  ) +
  theme(legend.position = "none") 
```

As you can see `flow` differs significantly based on `season`. Flow rates in Summer and spring are quite high, while those collected in winter and autumn are all less than 10 m^3/s. There is also a much smaller range of values for the Autumn and Winter flow values.


## Research Questions:

I am interested in exploring the following research questions:

1) What is the relationship between flow and month. I.e. does the month the flow rate was recorded effect flow?

2) Can we accurately predict flow from season alone? 

3) What effect does month have on extreme_type?

4) What are the long term trends in flow? Can we use this historical data to predict future flow patterns?

5) Does the historical data allow us to see a trend in flow that could be correlated with other climate data such as temperature and ice  

