---
title: "Data-Analysis-Milestone 2"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(datateachr)
library(tidyverse)
library(dplyr)
library(ggridges)
```

```{r}

flow_sample <- flow_sample %>% 
  mutate(season = case_when(
    month > 5 & month < 9 ~ "Summer",
    month > 2 & month < 6 ~ "Spring",
    (month > 0 & month < 3) | month == 12  ~ "Winter",
    month > 8 & month < 12 ~ "Autumn",
    TRUE ~ as.character(month)
  ) %>% as.factor()
  )

```

# Task 1

## Research Questions:

Here are the research questions I identified in milestone 1. I have reworked them slightly to better fit the tasks of this milestone:

1) What is the relationship between flow and season.

2) Can we accurately predict flow from month alone? 

3) What effect does extreme_type have on flow?

4) What is the relationship between year and flow?

## Task 1.2

### Exploring question 1

First I will compute the statistics for flow for the different seasons. First `Summer`
```{r}
flow_sample %>% filter(season == "Summer") %>% select(flow) %>% summary()
```
`Spring`

```{r}
flow_sample %>% filter(season == "Spring") %>% select(flow) %>% summary()
```

`Autumn`
```{r}
flow_sample %>% filter(season == "Autumn") %>% select(flow) %>% summary()
```
`Winter`

```{r}
flow_sample %>% filter(season == "Winter") %>% select(flow) %>% summary()
```


Now I find the mean of flow for the 4 seasons. .
```{r}
#mflow = ddply(flow_sample, "season", summarise, grp.mean=mean(flow))
#mflow
```
:( it was working before and then randomly stopped. I really don't know why. Sorry.

Finally I create a plot
```{r}
#plot1 <- ggplot(flow_sample, aes(x= flow, fill=season, color = season)) + geom_histogram(position="identity", alpha=0.5)
#plot1 +   geom_vline(data=mflow, aes(xintercept=grp.mean, color=season))
```

### Exploring question 2

I compute the range, mean, median, and sd across month
```{r}
flow_month_stats <- flow_sample %>%
  group_by(month) %>%
  summarise(mean = mean(flow), 
            r =range(flow), 
            sd = sd(flow), 
            var = var(flow))
```
I tried to create boxplots forwhat I thought would be the high and low flow months.
``` {r}
flow_sample <- flow_sample %>% 
  mutate(month_Flow = ifelse( month == 4 | month ==5 | month == 6 | month==7| month== 8, "High Flow", "Low Flow"))

flow_sample %>% 
  filter(!is.na(month)) %>% 
  ggplot() +
  aes(x=month, y=flow, fill= month) +
  geom_boxplot() +
  facet_wrap(~month_Flow, scales = "free") +
  labs(
    title = "Boxplot of flow by Season",
    x = "Season",
    y = "Flow"
  ) +
  theme(legend.position = "none") 
```

```{r}
flow_sample %>% 
  group_by(month) %>% 
  summarise(cumflow = sum(flow, na.rm = T)) %>% 
  ggplot() +
  aes(x=month, y=cumflow) +
geom_line() +
  geom_point() +
  scale_x_continuous(breaks = unique(flow_sample$month)) +
  labs(
    title = "Cumulative flow along the year",
  
    x = "Month",
    y = "Flow"
  ) 
```

### Exploring question 3
I compute the number of observations for `extreme_type` (categorical)
```{r}
flow_sample %>%
  count(extreme_type)
```
Now I graph extreme and flow
```{r}
flow_sample %>%
  ggplot(aes(x=flow, y=extreme_type)) +  
  geom_density_ridges(alpha=0.2, aes(fill= extreme_type))

```



### question 4
I will create a new (ordinal) categorical variable called decades. 

```{r}

flow_sample <- flow_sample %>% 
  mutate(decades = case_when(
    year > 1900 & year < 1930 ~ "1900-1929",
    year > 1929 & year < 1960 ~ "1930-1959",
    year > 1959 & year < 1990  ~ "1960-1989",
    year > 1989 & year < 2020 ~ "1990-2019",
    TRUE ~ as.character(year)
  ) %>% as.factor()
  )

```


```{r}
flow_sample %>% filter(decades == "1900-1929") %>% select(flow) %>% summary()
flow_sample %>% filter(decades == "1930-1959") %>% select(flow) %>% summary()
flow_sample %>% filter(decades == "1960-1989") %>% select(flow) %>% summary()
flow_sample %>% filter(decades == "1990-2019") %>% select(flow) %>% summary()

```

Now I will create a graph of flow and decades

```{r}
flow_sample %>% 
  group_by(decades) %>% 
   
  ggplot() +
  aes(x=decades, y=flow) +
  geom_line() +
  geom_point() +
  labs(
    title = "Flow within decades",
    x = "Decades",
    y = "Flow"
  )  
```

## Task 1.3

I am closer to answering my questions. Directly above I see that there is a lot of variation in flow within each 29 year period. However each of the bins seam to have a somewhat similar spread with perhaps a slight downward trend. The means do apear to be decreasing. There is the notable exception of the 2013 flood. 


There is obviously strong relationships between time of year (`month` and `season`), `extreme_type` and `flow`, but is one significantly better for predicting flow? 

Is the variation in flow year to year associated with the month in which flow was collected??

These are the questions I find most interesting.


# Task 2
Here I will tidy the data. Lets take a quick look at `flow_sample`. This is the before 
```{r}
head(flow_sample)
```

Given the definition I believe this data set is tidy. Each column is a variable `year`, `extreme_type` `month` `day`, `flow`, `sym` `season` and `decades`. Each row is an observation (one for each year). All of the cells contain the value for that variable on a given year.

## Task 2.1

Now I play with some functions from tidyverse.

I will unite the year, month, and day to make date.

```{r}
(flow_sample1 <- flow_sample %>%
    unite(col = "date", c(year, month, day), sep = "-"))
print(flow_sample1)
```
Now I separate them again.

```{r}
(flow_sample1.1 <- flow_sample1 %>%
    separate(date, into = c( "year", "month", "day"), sep = "-"))
print(flow_sample1.1)
```
I am now doing to untidy the data by widening the data so we see the flow of each `extreme_type` by  making `extreme_type` it's own column.

```{r}
(flow_sample2 <- flow_sample %>%
  pivot_wider(id_cols = c(-extreme_type, -flow),
              names_from = extreme_type,
              values_from = flow))
arrange_all(flow_sample2)
```
It is clear that there are many missing values. First I am going to check how many for maximum and how many for minimum.

```{r}
sum(is.na(flow_sample2$maximum))
sum(is.na(flow_sample2$minimum))
```
Most of these missing values are not an issue as only data on one extreme type was collected per year. I want to see the rows where both are NA.

```{r}
sum(is.na(flow_sample2$maximum) & is.na(flow_sample2$minimum) )

```
There are only 2 rows that have missing values for both extreme types. Since these rows have no flow information I am going to remove them. 

```{r}
which(is.na(flow_sample2$maximum) & is.na(flow_sample2$minimum) )

```
These are the two rows I will remove.

```{r}
flow_sample2.1 <- flow_sample2[-c(110,111),]
arrange_all(flow_sample2.1)
```
Now I will put it back. 

```{r}
flow_sample2.2 <- flow_sample2 %>% 
  pivot_longer(cols = c(maximum, minimum ), 
               names_to  = "extreme_type", 
               values_to = "flow")
arrange_all(flow_sample2.2)
```
You can see that we once again have variable `extreme_type` with values minimum and maximum.



## Task 2.3

```{r}
flow_sampleTR <-flow_sample[-c(110,111),] %>%
  filter(extreme_type == "maximum") %>%
select(-c(station_id, sym)) %>%
  select(year, season, flow, month, day, extreme_type)


arrange_all(flow_sampleTR)
```
This is the data I will carry forwards. I have removed the two row with no flow recorded. As well I removed `station_id` as there is only one station, and `sym` as there are more missing than present values. I filter for the extreme type `maximum` as there is much more variation and I think it may be more interesting to explore. Finally I rearrange the order of the columns.


